Tackling discrepancies in freshwater invertebrate trait databases: Harmonising across continents and aggregating taxonomic resolution

### 2021-02-19 Paper/Draft_harmonization_paper.tex
19.02: Suggestions Chuck

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\usepackage{float}
\usepackage[table,xcdraw]{xcolor}
\usepackage{color, colortbl}
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{tikz}
\usepackage{rotating}
\usepackage{caption}
\usepackage{authblk}
\usepackage{csquotes}
\graphicspath{{Figures/}}
\usepackage{setspace}
\usepackage{rotating}
\usepackage{geometry}
\usepackage{array}
\usepackage{lscape}
\usepackage{longtable}
\usepackage{hhline}
\usepackage{lmodern}
\usepackage[
backend = biber,
natbib,
style = apa6,
citestyle = authoryear,
maxcitenames = 2, 
maxbibnames = 99, 
uniquename=false,
uniquelist=false
%sorting = none % sort by name year title
]{biblatex}
\addbibresource{Ref_invertebrate_DB.bib}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Tackling discrepancies in freshwater invertebrate trait databases: Harmonising across continents and aggregating taxonomic resolution}
\author[1]{Stefan Kunz}
\author[2]{Ben J. Kefford}
\author[3]{Astrid Schmidt-Kloiber}
\author[4]{Christoph D. Matthaei}
\author[5]{Philippe Usseglio-Polatera}
\author[3]{Wolfram Graf}
\author[6]{N. LeRoy Poff}
\author[7]{Leon Metzeling}
\author[8]{Laura Twardochleb}
\author[9]{Charles P. Hawkins}
\author[1]{Ralf B. Schäfer}
\affil[1]{Institute for Environmental Sciences, University of Koblenz-Landau, Landau, Germany}
\affil[2]{Centre for Applied Water Science, Institute for Applied Ecology, University of Canberra, Canberra, Australia}
\affil[3]{Institute of Hydrobiology and Aquatic Ecosystem Management, University of Natural Resources and Life Sciences Vienna (BOKU), Vienna, Austria}
\affil[4]{Department of Zoology, University of Otago, Dunedin, New Zealand}
\affil[5]{University of Lorraine, CNRS, LIEC, Metz, France}
\affil[6]{Department of Biology, Colorado State University, Fort Collins, USA}
\affil[7]{Environment Protection Authority Victoria, Applied Sciences Division, Macleod, Australia}
\affil[8]{Department of Fisheries and Wildlife, Michigan State University, East Lansing, USA}
\affil[9]{Department of Watershed Sciences and the Ecology Center, Utah State University, Logan, USA}
\date{}

@@

1. Use of invertebrate traits rather than species composition may facilitate large-scale comparisons of community structure and responses to disturbance in freshwater ecology because the same traits can potentially occur everywhere.  In recent years, comprehensive invertebrate trait databases have been established at different scales (e.g. regions, continents). The wide availability of invertebrate trait data supports trait-based studies, especially at large scales. However, a number of data-related issues complicate the use of invertebrate traits for ecological studies. For example, standardised definitions for for freshwater invertebrate traits are lacking, which impedes comparisons across regions. Moreover, it is uncertain how harmonising varying trait definitions among databases might influence identification of trait-environment relationships. In addition, taxonomic resolution between observational taxonomic data and trait data can differ, making aggregation of traits necessary. At present, it is unknown how different trait aggregation approaches compare with expert-assigned trait affinities. In this study we aimed to identify discrepancies in trait definitions across freshwater invertebrate databases and to compare trait aggregation methods to identify how taxonomic hierarchy structure and trait variability influence results from different aggregation approaches.

@@

3. We found that among trait databases, trait definitions often differed because varying numbers of traits were used to describe groups of related traits (e.g., respiration traits) and the focus of descriptions for groups of related traits also varied (e.g., for feeding mode some databases focused on the food source, whereas others focused on mouthpart morphology). The coding to describe traits (binary, fuzzy) also varied among databases. Our comparison of different aggregation methods showed that family-level aggregated and expert-assigned traits were similar, especially from approaches using the median, and that trait coding method, uneven taxonomic hierarchical structure, and high trait variability can influence aggregated trait affinities. We further showed that harmonised and aggregated data performed similarly to observational data in identifying trait–environment relationships.

4. By outlining discrepancies in trait definitions we hope to motivate the development of standardised terminology for invertebrate traits. Our results also illustrate the usefulness of harmonised datasets for ecological study and provide guidance for the circumstances under which the choice of trait aggregation method is important.
\\
\\
\keywords{trait definitions, trait aggregation, data synthesis}

@@

Explaining and predicting how communities are shaped by environmental factors is a primary main goal of ecology. Species traits are measurable properties of an organism (\cite{mcgill_rebuilding_2006}), and comparing communities based on the traits possessed by species may facilitate testing of a range of ecological hypotheses (\cite{statzner_perspectives_2001}). Traits are assumed to be adaptations (e.g., physiological, behavioural) of organisms to their environment and to indicate direct or indirect linkages between the biological response of an organism or a population and its environment (\cite{southwood_habitat_1977, verberk_delivering_2013}). In addition to providing local-scale mechanistic-based expressions of species-environment relationships, trait-based approaches might be suitable for large-scale analyses because trait responses are less constrained by biogeographic boundaries and taxon distributional areas than taxonomic responses (\cite{baird_toward_2011, bonada_taxonomic_2007}).

Invertebrate traits have been increasingly used in freshwater ecology, e.g. by relating macroinvertebrate trait composition to environmental factors (\cite{bhowmik_large_2015, poff_developing_2010, szocs_effects_2014}). In the last decades, freshwater ecologists have compiled comprehensive invertebrate trait databases, which typically include data for a single region or continent (\cite{kefford_integrated_2020, Philips_and_Smith_NZ_DB_2018, schmidt-kloiber_www.freshwaterecology.info_2015, tomanova_trophic_2006, ussegliopolatera_biological_2000, vieira_database_nodate}). The availability of invertebrate trait data from different continents enables comparisons of regional trait variation and allows for testing the consistency of trait structure–environmental factor relationships across both small and large spatial extents. To date, such analyses have been carried out mostly within continents, using information from one or two trait databases. For example, \citet{bonada_taxonomic_2007} compared trait composition for Mediterranean and temperate regions in Europe based on traits from \citet{ussegliopolatera_biological_2000} (typically referred to as the Tachet database), \citet{poff_developing_2010} characterized trait composition across sites in the Western USA based on traits from \citet{poff_functional_2006}, and \citet{botwe_effects_2018} used trait definitions from \citet{poff_functional_2006} and \citet{schafer_trait_2011} to test for effects of salinity on invertebrate traits across different sites in South Australia. Analyses that synthesize invertebrate trait information from more than two different continents are rare, but see \citet{brown_functional_2018} and \citet{statzner_reproductive_1997}.

The heterogeneity of information in freshwater invertebrate trait databases, besides the diversity of taxa across regions, is likely a major reason for the lack of studies across continents. There are several challenges presented by using invertebrate trait data from multiple established databases in ecological studies. One such challenge is inconsistency in terminology. In this study, we follow the terminology proposed by \citet{schmera_proposed_2015}, where a grouping feature is defined as a general property (e.g. feeding mode) that comprises a "group of related traits (e.g., predator, shredder, etc.) that vary among species or among individuals within a species" (page number 825). Thus, to use consistent terminology in this study, we use the term grouping feature in place of the term trait, and the term trait instead of trait state, modality or trait category.

Another challenge to using data from multiple continents is that invertebrate trait databases include different grouping features and related traits. To harmonize grouping features from different regions, or to structure them in such a way as to facilitate cross-database comparison, commonly accepted and unambiguous trait definitions are required (\cite{schneider_towards_2019}). Ideally, the same traits and grouping features would be reported across databases or standardised terminology would exist to easily harmonise them; however, there is a lack of standardised terminology of trait definitions and poor metadata quality in many trait databases, making grouping-feature harmonisation difficult (\cite{baird_toward_2011, kissling_towards_2018}). To our knowledge, only \citet{brown_functional_2018} harmonised grouping features from more than two geographically distant invertebrate trait databases, for a limited set of grouping features (8) and taxa (112), in a study on the influence of decreasing glacier cover on functional diversity and invertebrate assemblages.

A third challenge to comparing across invertebrate trait databases is inconsistent coding of trait data (\cite{culp_incorporating_2011}), or assigning affinities (aka affinity scores). Traits of individual freshwater invertebrates can be difficult to quantify because, unlike plant traits, they are often difficult to measure directly. For example, to describe feeding habits requires that we understand mouthpart morphology, feeding behaviour, and specific food sources (\cite{moog_comprehensive_nodate}). Some traits can be described as binary or categorical, which ignores uncertainty in how traits are expressed in any particular organism (e.g., adult terrestrial stage, presence of gills). Other traits are better suited to being described as continuous (e.g., tolerance of pollution, body size). One approach for dealing with uncertainty is the use of fuzzy coding, where traits are assigned probabilistic values. Fuzzy codes, which are usually converted to proportions, are used to account for plasticity in traits, variability in traits within taxonomic groups above the species level, and incomplete knowledge. A challenge is presented by the use of different types of coding for trait affinities in different databases. For example, in the study from \citet{brown_functional_2018} the authors needed to reclassify trait affinities because the individual databases employed different coding approaches (i.e., European and New Zealand databases used fuzzy coding, North American database used binary coding).

Finally, discrepancies in the taxonomic resolutions (e.g., species, genus, or family) when comparing among trait databases, or when linking observational taxonomic data to trait databases, presents another challenge. When observations or data in one trait database are at a more precise taxonomic level than other data (e.g. observations at species-level and trait data at genus-level) trait information of the less precise taxonomic level is often assigned (e.g. \cite{szocs_effects_2014, vos_taxonomic_2017}). Conversely, if trait information is only available at more precise taxonomic levels than observed taxa, traits are aggregated to a less precise taxonomic level (e.g. \cite{aspin_extreme_2019, piliere_a._f._h._importance_2016, poff_functional_2006, szocs_effects_2014}). To date, studies have used different methods of trait aggregation, e.g. the mean (\cite{magliozzi_functional_2019}), median (\cite{szocs_effects_2014}) or mode (\cite{piliere_a._f._h._importance_2016}), but studies on how and to what extent different trait aggregation methods influence trait-based analyses are missing.

These challenges when working with or synthesizing trait data motivated us to assess the characteristics and comparability of freshwater invertebrate trait data contained in disparate databases. We aimed to explore methods of improving the comparability of trait data across databases, despite differences in trait definitions and the taxonomic level of data in three ways: (1) by investigating and describing discrepancies in trait definitions between trait databases from Europe, North America, Australia, and New Zealand; (2) by harmonising trait definitions across databases to provide consistent trait grouping features for comparison across geographic locations; and (3) by comparing the influence of different trait aggregation methods and grouping-feature harmonisation on the identification of trait–environment relationships.

@@

%Version from Brooke: As a basis for evaluating the comparability of freshwater invertebrate databases, we developed four novel invertebrate trait datasets by harmonising seven grouping features from existing trait databases. We then evaluated the influence of different...

To address our study objectives, we took a multi-prong approach to describing, combining, and assessing multiple sources of freshwater invertebrate data that differed in grouping-feature definitions and taxonomic resolution. Identified trait definitions discrepancies served as a starting point to develop four novel invertebrate trait datasets by harmonising seven grouping features from existing trait databases. The harmonised trait datasets were used to evaluate the influence of different trait aggregation approaches on the identification of trait–environment relationships, our third study objective, in two ways by (1) comparing family-level aggregated trait affinities with family-level trait affinities assigned by experts for two of the established trait datasets where such data were available and (2) comparing the influence of taxonomic hierarchy and trait variability on the outcomes of the different trait aggregation approaches for simulated data. Finally, to further investigate how harmonising and aggregating trait data can modify the outcomes of trait–environment relationship characterisation, we used data from one of the newly established datasets with harmonised and aggregated traits to repeat an analysis of salinisation effects on biological traits (\cite{szocs_effects_2014}) and compared our results with those of the original study.

@@

To explore the characteristics and comparability of freshwater invertebrate trait data with differing grouping-feature definitions and taxonomic resolution, we obtained invertebrate trait data from six trait databases for four regions: Europe, North America, Australia, and New Zealand. For Europe we obtained trait information from the freshwaterecology.info database (\cite{schmidt-kloiber_www.freshwaterecology.info_2015}), which contains taxa at the species level, and the Tachet database (\cite{ussegliopolatera_biological_2000}), which contains taxa at species, genus, and family levels. For North America we primarily obtained trait information from \citet{twardochleb_freshwater_nodate} (hereafter referred to as the CONUS database) and where taxa were missing from that database, we obtained data from \citet{vieira_database_nodate} (hereafter referred to as the Vieira database). Both databases contain taxa at species, genus and family levels. Data on body form for European and North American taxa were based on expert knowledge (\cite{polatera_personal_information_2020}). For Australia and New Zealand, we used trait databases from \citet{kefford_integrated_2020} (hereafter referred to as the Australian database) and \citet{Philips_and_Smith_NZ_DB_2018} (hereafter referred to as the New Zealand database) respectively. These databases contain taxa at species, genus, family, and less precise taxonomic levels.

We established four datasets, one for each of the four geographic regions, by harmonising the traits of seven grouping features from the trait databases. We selected traits that were available in all databases, are commonly applied in trait-based ecological studies, and describe different parts of the biology of a species: life history (Voltinism), morphology (Respiration, Body form, Size), ecology (Locomotion, Feeding mode) and reproduction (Oviposition). We omitted ecological traits that describe habitat preferences (e.g. temperature preference) because these traits are missing in the New Zealand trait database. Grouping features were  classified differently across the databases, so we harmonised them into 26 traits (Table \ref{tab:traits_harmonisation}). To do so we combined  similar traits into one trait (e.g. crawlers and sprawlers into crawlers) and assigned the highest trait affinity score among the combined traits for each taxon.

Next, we selected taxa for inclusion in our datasets and normalised the trait affinity scores for each taxon. We first omitted all taxa with a lower taxonomic precision than family level and consolidated duplicate taxa that appeared within a database, by either applying the median for fuzzy coded traits, or the maximum for binary traits. We then established our harmonised datasets by using either fuzzy coding or, when unavailable, binary coding. Fuzzy codes are reported with different ranges in the trait databases (e.g., freshwaterecology.info: 0–10, Tachet database: 0–3 or 0–5). We normalised these to a range between 0 and 1 and converted trait affinities to proportions. In cases where binary coding was needed, we converted categorical and continuous traits into binary traits with values of 1 and 0, which, thus, corresponded with the range of fuzzy coded traits. For example, in the  freshwaterecology.info database, the classification of the grouping feature voltinism accounts for different faunistic regions and we substituted a value of 1 for entries such as “arctic” or “boreal”. We assumed that values of 1 and 0 for binary traits corresponded to the highest and no affinity for a particular trait.

@@

For our study aim of comparing the effects of different aggregation methods on trait\-environment relationship inference, we aggregated the traits of two of the harmonised datasets to the family level.
We used three different aggregation approaches: (1) direct aggregation of taxa to family level with either the median or mean, giving equal weight to every taxon. We denote these aggregation methods as \textit{direct\_agg\textsubscript{median}} and \textit{direct\_agg\textsubscript{mean}}, respectively; (2) stepwise aggregation, first to the genus level and subsequently to the family level with either the median or mean. This approach gives equal weights to each genus. Hereafter, we denote this aggregation type as \textit{stepwise\_agg\textsubscript{median}} or \textit{stepwise\_agg\textsubscript{mean}}, respectively; and (3) aggregation using a weighted mean approach, denoted as \textit{weighted\_agg}. This method weights each genus according to its number of species in the trait datasets, regardless of whether information was available for every used grouping feature (Figure \ref{fig:data_proc_overview}). When we refer to \textit{direct\_agg\textsubscript{mean}}, \textit{stepwise\_agg\textsubscript{mean}}, and \textit{weighted\_agg} together we denote these methods as mean aggregation methods and when we refer to \textit{direct\_agg\textsubscript{median}} and \textit{stepwise\_agg\textsubscript{median}} together, we denote these methods as median aggregation methods.

@@

To evaluate the effect of the five trait aggregation methods
(\textit{direct\_agg\textsubscript{median}}, \textit{direct\_agg\textsubscript{mean}}, \textit{stepwise\_agg\textsubscript{median}}, \textit{stepwise\_agg\textsubscript{mean}}, \textit{weighted\_agg}) on the assignment of trait affinity scores, we compared our aggregated affinities with trait affinities assigned at family level by experts. Expert-assigned trait data were available for a subset of grouping features and taxa for Australia (\cite{chessman_dissolved-oxygen_2018}) and North America (CONUS database), so we conducted this comparison only for these two datasets. For the Australian dataset, we used the grouping features feeding mode (all traits listed in Table 1 except parasite) and size (220 families $*$ 8 family–trait combinations = 1760 comparisons). For the North American dataset, we used the grouping features feeding mode, respiration, size, voltinism, and locomotion (94 families $*$ 17 family–trait combinations = 1598 comparisons). Assigned traits from North America were on a categorical scale, and we converted them to binary prior to comparing them with aggregated trait affinities. Size was assigned as continuous variable for Australia, thus, we converted size into three traits (small, medium, and large) with a binary coding. Because trait affinities ranged from 0 to 1, the maximum difference possible in trait affinity was 1 to –1, either which corresponds to 100 \%. For convenience and to improve interpretation, we calculated absolute trait differences.

@@

\subsubsection*{Effect of taxonomic hierarchy and trait variability on aggregation outcomes}

>>>>

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\usepackage{float}
\usepackage[table,xcdraw]{xcolor}
\usepackage{color, colortbl}
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{tikz}
\usepackage{rotating}
\usepackage{caption}
\usepackage{ccaption}
\usepackage{authblk}
\usepackage{csquotes}
\graphicspath{{Figures/}}
\usepackage{setspace}
\usepackage{rotating}
\usepackage{geometry}
\usepackage{array}
\usepackage{lscape}
\usepackage{longtable}
\usepackage{hhline}
\usepackage{lmodern}
\usepackage[
backend = biber,
natbib,
style = apa6,
citestyle = authoryear,
maxcitenames = 2, 
maxbibnames = 99, 
uniquename=false,
uniquelist=false
%sorting = none % sort by name year title
]{biblatex}
\addbibresource{Ref_invertebrate_DB.bib}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Tackling discrepancies in freshwater invertebrate trait databases: Harmonising across continents and aggregating taxonomic resolution}
\author[1]{Stefan Kunz}
\author[2]{Ben J. Kefford}
\author[3]{Astrid Schmidt-Kloiber}
\author[4]{Christoph D. Matthaei}
\author[5]{Philippe Usseglio-Polatera}
\author[3]{Wolfram Graf}
\author[6]{N. LeRoy Poff}
\author[7]{Leon Metzeling}
\author[8]{Laura Twardochleb}
\author[9]{Charles P. Hawkins}
\author[1]{Ralf B. Schäfer}
\affil[1]{Institute for Environmental Sciences, University of Koblenz-Landau, Landau, Germany}
\affil[2]{Centre for Applied Water Science, Institute for Applied Ecology, University of Canberra, Canberra, Australia}
\affil[3]{Institute of Hydrobiology and Aquatic Ecosystem Management, University of Natural Resources and Life Sciences Vienna (BOKU), Vienna, Austria}
\affil[4]{Department of Zoology, University of Otago, Dunedin, New Zealand}
\affil[5]{University of Lorraine, CNRS, LIEC, Metz, France}
\affil[6]{Department of Biology, Colorado State University, Fort Collins, USA}
\affil[7]{Environment Protection Authority Victoria, Applied Sciences Division, Macleod, Australia}
\affil[8]{Department of Fisheries and Wildlife, Michigan State University, East Lansing, USA}
\affil[9]{Department of Watershed Sciences, National Aquatic Monitoring Center, and the Ecology Center, Utah State University, Logan, USA}
\date{}

@@

1. Use of invertebrate traits rather than species composition may facilitate large-scale comparisons of community structure and responses to disturbance in freshwater ecology because the same traits can potentially occur everywhere.  In recent years, comprehensive invertebrate trait databases have been established at different scales (e.g. regions, continents). The wide availability of invertebrate trait data supports trait-based studies, especially at large scales. However, a number of data-related issues complicate the use of invertebrate traits for ecological studies. For example, standardised definitions for freshwater invertebrate traits are lacking, which impedes comparisons across regions. Moreover, it is uncertain how harmonising varying trait definitions among databases might influence identification of trait-environment relationships. In addition, taxonomic resolution between taxonomic composition data and trait data can differ, making aggregation of traits necessary. At present, it is unknown how different trait aggregation approaches compare with expert-assigned trait affinities. In this study we aimed to identify discrepancies in trait definitions across freshwater invertebrate databases and to compare trait aggregation methods to identify how taxonomic hierarchy structure and trait variability influence results from different aggregation approaches.

@@

3. We found that among trait databases, trait definitions often differed because varying numbers of traits were used to describe groups of related traits (e.g., respiration traits) and the focus of descriptions for groups of related traits also varied (e.g., for feeding mode some databases focused on the food source, whereas others focused on mouthpart morphology). The coding to describe traits (binary, fuzzy) also varied among databases. Our comparison of different aggregation methods showed that family-level aggregated and expert-assigned traits were similar, especially when aggregation was based on the median, and that trait coding method, uneven taxonomic hierarchical structure, and high trait variability can influence aggregated trait affinities. We further showed that harmonised and aggregated data performed similarly to not aggregated and harmonised data from the case study in identifying trait–environment relationships.

4. By identifying discrepancies in trait definitions we hope to motivate the development of standardised terminology for invertebrate traits. Our results also illustrate the usefulness of harmonised datasets for ecological study and provide guidance for the circumstances under which the choice of trait aggregation method is important.
\\
\\
\keywords{trait definitions, trait aggregation, data synthesis}

@@

Explaining and predicting how communities are shaped by environmental factors is a primary goal of ecology. Species traits are measurable properties of an organism (\cite{mcgill_rebuilding_2006}), and comparing communities based on the traits possessed by species may facilitate testing of a range of ecological hypotheses (\cite{statzner_perspectives_2001}). Traits are assumed to be adaptations (e.g., physiological, behavioural) of organisms to their environment, which represent either direct or indirect linkages between the biological response of an organism or a population and its environment (\cite{southwood_habitat_1977, verberk_delivering_2013}). In addition to providing local-scale mechanistic-based expressions of species-environment relationships, trait-based approaches might be suitable for large-scale analyses of environmental risks to communities because the occurrence and distribution of traits are less constrained by biogeographic boundaries than that of taxa (\cite{baird_toward_2011, bonada_taxonomic_2007}).

Invertebrate traits have been increasingly used in freshwater ecology, e.g. by relating macroinvertebrate trait composition to environmental factors such as climate change or salinisation (\cite{bhowmik_large_2015, poff_developing_2010, szocs_effects_2014}). In the last decades, freshwater ecologists have compiled comprehensive invertebrate trait databases, which typically include data for a single region or continent (\cite{kefford_integrated_2020, Philips_and_Smith_NZ_DB_2018, schmidt-kloiber_www.freshwaterecology.info_2015, tomanova_trophic_2006, ussegliopolatera_biological_2000, vieira_database_nodate}). The availability of invertebrate trait data from different continents enables comparisons of regional trait variation and allows for testing the consistency of trait structure–environmental factor relationships across both small and large spatial extents. To date, such analyses have been carried out mostly within continents, using information from one or two trait databases. For example, \citet{bonada_taxonomic_2007} compared trait composition for Mediterranean and temperate regions in Europe based on traits from \citet{ussegliopolatera_biological_2000} (typically referred to as the Tachet database), \citet{poff_developing_2010} characterized trait composition across sites in the Western USA based on traits from \citet{poff_functional_2006}, and \citet{botwe_effects_2018} used trait definitions from \citet{poff_functional_2006} and \citet{schafer_trait_2011} to test for effects of salinity on invertebrate traits across different sites in South Australia. Analyses that synthesize invertebrate trait information from more than two different continents are rare, but see \citet{brown_functional_2018} and \citet{statzner_reproductive_1997}.

The variability and inconsistencies of information in freshwater invertebrate trait databases, besides the diversity of taxa across regions, is likely a major reason for the lack of comparisons across continents. Several challenges currently prevent use of invertebrate trait data from different established databases in ecological studies. One challenge is inconsistency in terminology. In this study, we follow the terminology proposed by \citet{schmera_proposed_2015}, in which a grouping feature (e.g. feeding mode) refers to a set of related organismal traits (e.g. predator, shredder, etc.) that together define an ecological property that is shared among individuals or species. Thus, to use consistent terminology in this study, we use the term grouping feature in place of the term trait, and the term trait instead of trait state, modality or trait category.

Another challenge to using data from multiple continents is that invertebrate trait databases include different grouping features and related traits. To harmonise grouping features from different regions, or to structure them in such a way as to facilitate cross-database comparison, commonly accepted and unambiguous trait definitions are required (\cite{schneider_towards_2019}). Ideally, the same traits and grouping features would be reported across databases or standardised terminology would exist to easily harmonise them; however, standardised terminology of trait definitions is lacking and poor metadata quality in many trait databases make, harmonisation of grouping features difficult (\cite{baird_toward_2011, kissling_towards_2018}). To our knowledge, only \citet{brown_functional_2018} harmonised grouping features from more than two geographically distant invertebrate trait databases, for a limited set of grouping features (8) and taxa (112), in a study on the influence of decreasing glacier cover on the functional diversity of invertebrate assemblages.

A third challenge to comparing across invertebrate trait databases is inconsistent coding of trait data (\cite{culp_incorporating_2011}). Coding of trait data is usually done by assigning an affinity (aka affinity score) of an invertebrate taxon to express a particular trait %, either using a binary classification (0/1), continuous or fuzzy classification (0 - 1 or 0 - 100 \%).
Traits of individual freshwater invertebrates can be difficult to quantify because, unlike plant traits, they are often difficult to measure directly. For example, to describe feeding habits requires that we understand mouthpart morphology, feeding behaviour, and what is being eaten (\cite{moog_comprehensive_nodate}). Some traits can be described in categorical terms, which ignores uncertainty in how traits are expressed in any particular organism (e.g., adult terrestrial stage, presence of gills). Other traits are better suited to being described as continuous (e.g., tolerance of pollution, body size). One approach for dealing with uncertainty is the use of fuzzy coding, where traits are assigned probabilistic values. Fuzzy codes, which are usually converted to proportions, are used to account for plasticity in traits, variability in traits within taxonomic groups above the species level, and incomplete knowledge. The use of different types of coding for trait affinities in different databases limits comparisons. For example, in the study from \citet{brown_functional_2018} the authors needed to reclassify trait affinities because the individual databases employed different coding approaches (i.e., European and New Zealand databases used fuzzy coding whereas the North American database used binary coding).

Finally, discrepancies in the taxonomic resolutions (e.g., species, genus, or family) used when comparing among trait databases, or when linking counted taxonomic data to trait databases, presents another challenge. When taxonomic composition or data in on trait database are at a more precise taxonomic level than other trait data (e.g. observations at species-level and trait data at genus-level) trait information of the less precise taxonomic level is often assigned (e.g. \cite{szocs_effects_2014, vos_taxonomic_2017}). If trait information is only available at more precise taxonomic levels than for taxonomic composition, traits are aggregated to a less precise taxonomic level (e.g. \cite{aspin_extreme_2019, piliere_a._f._h._importance_2016, poff_functional_2006, szocs_effects_2014}). To date, studies have used different methods of trait aggregation, e.g. the mean (\cite{magliozzi_functional_2019}), median (\cite{szocs_effects_2014}) or mode (\cite{piliere_a._f._h._importance_2016}), but studies on how and to what extent different trait aggregation methods influence trait-based analyses are missing.

These challenges motivated us to assess the characteristics and comparability of different freshwater invertebrate trait databases with the goal of facilitating future syntheses. We aimed to identify methods of improving the comparability of trait data across databases, despite differences in trait definitions and the taxonomic level of data in three ways: (1) by describing discrepancies in trait definitions between trait databases from Europe, North America, Australia, and New Zealand; (2) by harmonising trait definitions across databases to provide consistent trait grouping features for comparison across geographic locations; and (3) by comparing the influence of different trait aggregation methods and grouping-feature harmonisation on the detection of trait–environment relationships.

@@

To address our study objectives, we took a multi-prong approach to describing, combining, and assessing multiple sources of freshwater invertebrate data that differed in grouping-feature definitions and taxonomic resolution. Discrepancies in trait definitions served as a starting point to develop four novel invertebrate trait datasets by harmonising seven grouping features present in the existing trait databases. The harmonised trait datasets were used to evaluate the influence of different trait aggregation approaches on the identification of trait–environment relationships, our third study objective, in two ways by (1) comparing family-level aggregated trait affinities with family-level trait affinities assigned by experts for two of the established trait datasets where such data were available and (2) comparing the influence of taxonomic hierarchy and trait variability on the outcomes of the different trait aggregation approaches for simulated data. Finally, to further investigate how harmonising and aggregating trait data can modify the outcomes of trait–environment relationship characterisation, we used data from one of the newly established datasets with harmonised and aggregated traits to repeat an analysis of salinisation effects on biological traits (\cite{szocs_effects_2014}) and compared our results with those of the original study.

@@

We assessed how six databases differed in how they assigned and described traits of freshwater invertebrates. These databases (from Europe, North America, Australia, and New Zealand) differed in how grouping features were defined and the taxonomic resolution used to assign traits. For Europe we obtained trait information from the freshwaterecology.info database (\cite{schmidt-kloiber_www.freshwaterecology.info_2015}), which assigns traits at the species level, and the Tachet database (\cite{ussegliopolatera_biological_2000}), which assigns traits at species, genus, and family levels. For North America we primarily obtained trait information from \citet{twardochleb_freshwater_nodate} (hereafter referred to as the CONUS database) and where taxa were missing from that database, we obtained data from \citet{vieira_database_nodate} (hereafter referred to as the Vieira database). Both databases contain taxa at species, genus and family levels. Data on body form for European and North American taxa were based on expert knowledge (\cite{polatera_personal_information_2020}). For Australia and New Zealand, we used trait databases from \citet{kefford_integrated_2020} (hereafter referred to as the Australian database) and \citet{Philips_and_Smith_NZ_DB_2018} (hereafter referred to as the New Zealand database). These databases assigned traits at species, genus, family, and coarser taxonomic levels.

We established four datasets, one for each of the four geographic regions, by harmonising the traits of seven grouping features from the trait databases. We selected traits that were available in all databases, are commonly applied in trait-based ecological studies, and describe different aspects of the biology of a species: life history (Voltinism), morphology (Respiration, Body form, Size), ecology (Locomotion, Feeding mode) and reproduction (Oviposition). We omitted ecological traits that describe habitat preferences (e.g. temperature preference) because these traits are missing in the New Zealand trait database. Grouping features were classified differently across the databases, so we harmonised 7 grouping features that collectively included 26 traits (Table \ref{tab:traits_harmonisation}). To do so we combined similar traits into one trait (e.g. crawlers and sprawlers into crawlers) and for each taxon we assigned for the new combined trait the highest affinity score of those originally assigned to the subsequently combined traits. 
% and assigned the highest trait affinity score among the combined traits for each taxon.  
Next, we selected taxa for inclusion in our datasets and normalised the trait affinity scores for each taxon. We first omitted all taxa with a lower taxonomic precision than family level and consolidated duplicate taxa that appeared within a database, by either applying the median for fuzzy coded traits, or the maximum for binary traits. We then used either fuzzy coding or, when unavailable, binary coding to create the final harmonised datasets. Fuzzy codes are reported with different ranges in the trait databases (e.g., freshwaterecology.info: 0–10, Tachet database: 0–3 or 0–5). We normalised these to a range between 0 and 1 and converted trait affinities to proportions. In cases where binary coding was needed, we converted categorical and continuous traits into binary traits with values of 1 and 0, which, therefore, had the same range as fuzzy coded traits. For example, in the  freshwaterecology.info database, voltinism traits are associated with different faunistic regions, and we substituted a value of 1 for entries such as “arctic” or “boreal” % and 0 if no region was given. 
We assumed that values of 1 and 0 for binary traits corresponded to the highest and lowest affinity for a particular trait.

@@

To compare the effects of different aggregation methods on trait\-environment relationship inferences, we aggregated the traits of two of the harmonised datasets to the family level. We used three different aggregation approaches: (1) direct aggregation of taxa to family level, in which we aggregated either trait affinities by the median or mean for species and genera within a family, giving equal weight to every taxon. We denote these aggregation methods as \textit{direct\_agg\textsubscript{median}} and \textit{direct\_agg\textsubscript{mean}}, respectively; and (2) stepwise aggregation, first to the genus level and subsequently to the family level with either the median or mean. This approach gives equal weights to each genus. Hereafter, we denote this aggregation type as \textit{stepwise\_agg\textsubscript{median}} or \textit{stepwise\_agg\textsubscript{mean}}, respectively; and (3) aggregation using a weighted mean approach, denoted as \textit{weighted\_agg}. This method weights each genus according to its number of species in the trait datasets, regardless of whether information was available for every used grouping feature (Figure \ref{fig:data_proc_overview}). When we refer to \textit{direct\_agg\textsubscript{mean}}, \textit{stepwise\_agg\textsubscript{mean}}, and \textit{weighted\_agg} together we denote these methods as mean aggregation methods and when we refer to \textit{direct\_agg\textsubscript{median}} and \textit{stepwise\_agg\textsubscript{median}} together, we denote these methods as median aggregation methods.

@@

To evaluate the effect of the five trait aggregation methods
(\textit{direct\_agg\textsubscript{median}}, \textit{direct\_agg\textsubscript{mean}}, \textit{stepwise\_agg\textsubscript{median}}, \textit{stepwise\_agg\textsubscript{mean}}, \textit{weighted\_agg}) on the assignment of trait affinity scores, we compared our aggregated affinities with trait affinities assigned at family level by experts. Expert-assigned trait data were available for a subset of grouping features and taxa for Australia (\cite{chessman_dissolved-oxygen_2018}) and North America (CONUS database), so we conducted this comparison only for these two datasets. For the Australian dataset, we used the grouping features feeding mode (all traits listed in Table 1 except parasite) and size (220 families $\times$ 8 family–trait combinations = 1760 comparisons). For the North American dataset, we used the grouping features feeding mode, respiration, size, voltinism, and locomotion (94 families $\times$ 17 family–trait combinations = 1598 comparisons). Assigned traits from North America were on a categorical scale, and we converted them to binary prior to comparing them with aggregated trait affinities. Size is a continuous variable in the Australian database, so, we converted size into three traits (small, medium, and large) each with a binary coding. Because trait affinities ranged from 0 to 1, the maximum difference possible in trait affinity was 1 to –1, both of which corresponds to 100 \%. For convenience and to improve interpretation, we calculated absolute values of the trait differences, i.e. a difference of -1 was converted to 1.

@@

\subsubsection*{Effect of phylogeny and trait variability on aggregation outcomes}
